---
title: "9131 Rutland Road assessment analysis"
format:
  html:
    theme: flatly
execute: 
  echo: true
  warning: false
  freeze: auto
code-fold: true
code-overflow: scroll
editor: visual
---

```{r}
#| label: setup
#| code-summary: "Setup code"

library(tidyverse)
library(hdatools)
library(scales)

setwd("~/repos/sandbox/rutland")

```

Get land and improvement values for parcels from Hanover County [online GIS](https://parcelmap.hanovercounty.gov/).

Parcels included:

-   9131 Rutland Road (focus parcel)

-   9125 Rutland Road (adjoining south)

-   9135 Rutland Road (adjoining north)

-   9143 Rutland Road (adjoins 9135 to north)

```{r}
#| label: import
#| code-summary: "Import data"

assessments <- read_csv("assessments.csv") |> 
  
  # Add total assessed value column
  
  mutate(total = land + improvement) |> 

  # Filter out 2017 and earlier
  
  filter(year > 2017) |> 
  
  # Rearrange
  
  pivot_longer(
    cols = 4:6,
    names_to = "type",
    values_to = "value"
  )

```

```{r}
#| label: data
#| code-summary: "Arrange data"

assessments_pct <- assessments |> 
  group_by(parcel, type) |> 
  arrange(year) |> 
  mutate(chg = value - first(value),
         pct_chg = chg / first(value)) |> 
  ungroup() |> 
  mutate(parcel = fct_reorder(parcel, pct_chg),
         alpha = case_when(
           parcel == "9131 Rutland Road" ~ "1",
           TRUE ~ "0"
         ))

```

## base_size not specified

`hdatools::theme_hda()` default base size is 14.

```{r}
#| label: plot-1
#| code-fold: false

ggplot(assessments_pct, aes(x = year, y = pct_chg, color = parcel, alpha = alpha)) +
  geom_line(linewidth = 1.5) +
  facet_wrap(~type, nrow = 1) +
  scale_y_continuous(label = label_percent()) +
  scale_color_hda(direction = -1) +
  scale_alpha_manual(values = c(0.5, 1), guide = FALSE) +
  guides(color = guide_legend(reverse = TRUE)) +
  labs(title = "Percent change in assessed value",
       subtitle = "2018 through 2023 (proposed)") +
  theme_hda() +
  add_zero_line("y") +
  theme(
    legend.position = "top",
    axis.text.x = element_text(angle = 90)
  )

```

## base_size adjusted

Need to adjust base size to 23 to scale properly.

```{r}
#| label: plot-2
#| code-fold: false

ggplot(assessments_pct, aes(x = year, y = pct_chg, color = parcel, alpha = alpha)) +
  geom_line(linewidth = 1.5) +
  facet_wrap(~type, nrow = 1) +
  scale_y_continuous(label = label_percent()) +
  scale_color_hda(direction = -1) +
  scale_alpha_manual(values = c(0.5, 1), guide = FALSE) +
  guides(color = guide_legend(reverse = TRUE)) +
  labs(title = "Percent change in assessed value",
       subtitle = "2018 through 2023 (proposed)") +
  theme_hda(base_size = 23) +
  add_zero_line("y") +
  theme(
    legend.position = "top",
    axis.text.x = element_text(angle = 90)
  )

```

## Without using hdatools functions

Base size similarly incorrect even when `hdatools` functions are not used.

```{r}
#| label: plot-3
#| code-fold: false

ggplot(assessments_pct, aes(x = year, y = pct_chg, color = parcel, alpha = alpha)) +
  geom_line(linewidth = 1.5) +
  facet_wrap(~type, nrow = 1) +
  scale_y_continuous(label = label_percent()) +
#  scale_color_hda(direction = -1) +
  scale_alpha_manual(values = c(0.5, 1), guide = FALSE) +
  guides(color = guide_legend(reverse = TRUE)) +
  labs(title = "Percent change in assessed value",
       subtitle = "2018 through 2023 (proposed)") +
#  theme_hda(base_size = 23) +
#  add_zero_line("y") +
  theme(
    legend.position = "top",
    axis.text.x = element_text(angle = 90)
  )

```
