---
title: "fmr"
format: html
---

```{r}

library(tidyverse)
library(readxl)
library(scales)

library(showtext)

# font_add_google(name = "Open Sans", family = "Open Sans")
# 
# showtext_auto()

```

## Import data

Import FMR files for FY 2022 and FY 2023

```{r}

fmr_2022_raw <- read_xlsx("data/FY22_FMRs_revised.xlsx")

fmr_2023_raw <- read_xlsx("data/FY23_FMRs.xlsx")

```

## Wrangle data

Filter to Virginia, remove unnecessary columns, and add columns to denote applicable fiscal year. Collapse records to remove duplicate entries for localities within a single FMR area (i.e., MSA).

```{r}

fmr_2022 <- fmr_2022_raw |> 
  filter(state == "51") |> 
  select(metro_code, areaname, fmr_2022 = fmr_2) |> 
  distinct(metro_code, .keep_all = TRUE)

fmr_2023 <- fmr_2023_raw |> 
  filter(state == "51") |> 
  select(metro_code = hud_area_code, areaname = hud_area_name, fmr_2023 = fmr_2) |> 
  distinct(metro_code, .keep_all = TRUE)

```

Join data and calculate percent change from FY 2022 to FY 2023. Filter for only metro regions.

```{r}

fmrs <- fmr_2022 |> 
  left_join(select(fmr_2023, fmr_2023, metro_code), by = "metro_code") |> 
  mutate(pct_chg = ((fmr_2023 - fmr_2022)/fmr_2022)) |> 
  filter(str_detect(metro_code, "METRO")) |> 
  mutate(areaname = str_remove_all(areaname, ", VA"),
         areaname = str_remove_all(areaname, "HUD Metro ")) |> 
  na.omit()

```

## Plot data

Create ordered lollipop chart of percent change in FMRs from FY 2022 to FY 2023.

```{r}

ggplot(fmrs,
       aes(x = pct_chg,
           y = reorder(areaname, pct_chg))) +
  geom_col(width = 0.175,
           fill = "#e8977a") +
  geom_point(size = 1.5,
             color = "#e8977a") +
  scale_x_continuous(labels = label_percent(),
                     expand = expansion(mult = c(0,0.05))) +
  labs(title = "Percent change in 2-bedroom Fair Market Rent: FY 2022 to FY 2023",
       caption = "Source: U.S. Department of Housing and Urban Development.") +
  theme_minimal(base_size = 18) +
  theme(panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank(),
        panel.grid.major.y = element_line(size = 0.15),
        axis.title = element_blank(),
        plot.title.position = "plot",
        )

ggsave("fmr_plot.png", width = 1200, height = 800, units = "px", bg = "#FFFFFF")

```

